\Large {\bf rt1d: Documentation}

### 2012-05-23 doc/manual/rt1d_manual.tex
Minor changes. About to rework lookup table initialization and interpolation to accomodate higher (4-5) dimensional tables.

\usepackage{amsmath, amsthm, amssymb}
\numberwithin{equation}{section}
\usepackage{graphicx}
\usepackage{pslatex}
\usepackage{epsfig}
\usepackage{natbib}
\usepackage[pdfborder={0 0 1}]{hyperref}

@@

\setcounter{tocdepth}{2}
\tableofcontents
\newpage

@@

% Rate Equations
\subsection{The Rate Equations}
In general, the chemical and thermal evolution of gas surrounding a radiation
source is governed by a set of differential equations describing the number
densities of all ions and the temperature of the gas. Assuming a medium
consisting of hydrogen and helium only, we first solve for the abundances of
each ion via
\begin{align}
    \frac{d \nHII}{dt} & = (\gHI + \gamma_{\HI} + \bHI \nel) \nHI - \aHII \nel \nHII   \label{eq:HIIRateEquation} \\ 
    \frac{d \nHeII}{dt} & = (\gHeI + \gamma_{\HeI} + \bHeI \nel) \nHeI \nonumber + \aHeIII \nel \nHeIII \\  & - (\bHeII + \aHeII + \xiHeII) \nel \nHeII \label{eq:HeIIRateEquation} \\ 
    \frac{d \nHeIII}{dt} & = (\gHeII + \gamma_{\HeII} + \bHeII \nel) \nHeII  - \aHeIII \nel \nHeIII . \label{eq:HeIIIRateEquation}
\end{align}
Each of these equations represents the balance between ionizations of species
$\HI$, $\HeI$, and $\HeII$, and recombinations of $\HII$, $\HeII$, and
$\HeIII$. Associating the index $i$ with absorbing species, $i = \HI, \HeI,
\HeII$, we define $\Gamma_i$ as the photo-ionization rate coefficient,
$\gamma_i$ as the secondary ionization rate coefficient, $\alpha_i$ ($\xi_i$)
as the case-B (dielectric) recombination rate coefficients, $\beta_i$ as the
collisional ionization rate coefficients, and $\nel = \nHII + \nHeII +
2\nHeIII$ as the number density of electrons.

At each time-step, we also solve for the temperature evolution, $dT_k/dt$,
which
is given by
\begin{align}
    \frac{3}{2}\frac{d}{dt}\left(\frac{\kB T_k \ntot}{\mu}\right) & = \fheat  \sum_i n_i \Heat_i - \sum_i \zeta_i n_e n_i - \sum_{i^{\prime}} \eta_i n_e n_i \nonumber \\ & - \sum_i \psi_i n_e n_i - \omega_{\HeII} n_e \nHeII \label{eq:TemperatureEvolution} 
\end{align}
where the index $i$ corresponds as usual to species that have bound electrons,
and $i^{\prime}$ corresponds to ionized species. $\Heat_i$ is the
photo-electric heating rate coefficient (due to electrons previously bound to
species $i$), $\omega_{\HeII}$ is the dielectric recombination cooling
coefficient, and $\zeta_i$, $\eta_i$, and $\psi_i$ are the collisional
ionization, recombination, and collisional excitation cooling coefficients,
respectively. The constants in Equation \ref{eq:TemperatureEvolution} are the
total number density of baryons, $\ntot = n_\mathrm{H} + n_{\mathrm{He}} +
n_e$, the mean molecular weight, $\mu$, Boltzmann's constant, $\kB$, and the
fraction of secondary electron energy deposited as heat, $\fheat$. We use the
formulae in Appendix B of \citet{Fukugita1994} to compute the values of
$\alpha_i$, $\beta_i$, $\xi_i$, $\zeta_i$, $\eta_i$, $\psi_i$, and
$\omega_{\HeII}$.

% Ionization and Heating Rates
\subsection{Computing the Ionization and Heating Rates}
The most critical aspect of propagating the radiation field in our 1D
simulations is computing the ionization ($\Gamma_i$, $\gamma_{i}$) and heating
($\Heat_i$) rate coefficients accurately. In order to directly relate our
results to fully 3-dimensional radiative transfer calculations we have chosen
to adopt a photon-conserving (PC) algorithm nearly identical to those employed
by several widely used codes, like \textit{C$^2$Ray} \citep{Mellema2006}, and
\textit{Enzo} \citep{Wise2011}. Our code is able to compute $\Gamma_i$,
$\gamma_{i}$, and $\Heat_i$ in a non-photon-conserving (NPC) fashion as well,
to enable comparison with previous 1D work such as \citet{Thomas2008}. The two
formalisms are equivalent in the limit of very optically thin cells, a
condition that can be met easily in 1D calculations but is rarely
computationally feasible in 3D. For NPC methods, if the optical depth of an
individual cell is substantial, the number of ionizations in that cell will
\textit{not} equal the number of photons absorbed for that cell, i.e. photon
number will not be conserved. This problem was remedied by \citet{Abel1999},
who inferred the number of photo-ionizations of species $i$ in a cell from the
radiation incident upon it and its optical depth,
\begin{equation}
    \Delta \tau_{i,\nu} = n_i \sigma_{i,\nu} \Delta r .
\end{equation}    
It is most straightforward to imagine our 1D grid as a collection of
concentric spherical shells, each having thickness $\Delta r$ and volume
$V_{\sh}(r) = 4 \pi [(r + \Delta r)^3 - r^3] / 3$ (i.e. radii correspond to
cell interfaces). The ionization and heating rates can then be related to the
number of absorptions in any given shell (thus preserving photon number), as
\begin{align}
    \Gamma_i & = A_i \int_{\nu_i}^{\infty} I_{\nu} e^{-\tau_{\nu}} \left(1 - e^{-\Delta \tau_{i,\nu}}\right) \frac{d\nu}{h\nu} \label{eq:PhotoIonizationRate} \\
    \gamma_{ij} & = A_j \int_{\nu_j}^{\infty} \left(\frac{\nu - \nu_j}{\nu_i}\right) I_{\nu} e^{-\tau_{\nu}} \left(1 - e^{-\Delta \tau_{j,\nu}}\right) \frac{d\nu}{h\nu} \label{eq:SecondaryIonizationRate} \\
    \Heat_i & = A_i \int_{\nu_i}^{\infty} (\nu - \nu_i) I_{\nu} e^{-\tau_{\nu}} \left(1 - e^{-\Delta \tau_{i,\nu}}\right) \frac{d\nu}{\nu} , \label{eq:HeatingRate}
\end{align}    
where we've defined the normalization constant, $A_i \equiv
L_{\mathrm{bol}}/n_i V_{\sh}(r)$, and denote the ionization threshold energy
for species $i$ as $h\nu_i$.

@@

@@

@@

%%%
%% SED studies
%%%
\section{Verifying Discrete SEDs with \texttt{rt1d}}
In \citet{Mirocha2012}, we used \texttt{rt1d} to verify new, optimally selected discrete spectral energy distributions for sources of ultraviolet and X-ray photons.  As we optimize more and more SEDs for fully 3D cosmological  radiation-hydro simulations, \texttt{rt1d} remains a quick and easy tool for verifying that the SEDs we're using are acceptable.

>>>>

\usepackage{amsmath, amsthm, amssymb}
\numberwithin{equation}{section}
\usepackage{graphicx}
\usepackage{pslatex}
\usepackage{epsfig}
\usepackage{natbib}
\usepackage[pdfborder={0 0 1}]{hyperref}
\usepackage[htt]{hyphenat}

@@

% Table of contents
\setcounter{tocdepth}{2}
\tableofcontents
\newpage

@@

% Rate Equations
\subsection{The Rate Equations}
In general, the chemical and thermal evolution of gas surrounding a radiation
source is governed by a set of differential equations describing the number
densities of all ions and the temperature of the gas. Assuming a medium
consisting of hydrogen and helium only, we first solve for the abundances of
each ion via
\begin{align}
    \frac{d \nHII}{dt} & = (\gHI + \gamma_{\HI} + \bHI \nel) \nHI - \aHII \nel \nHII   \label{eq:HIIRateEquation} \\ 
    \frac{d \nHeII}{dt} & = (\gHeI + \gamma_{\HeI} + \bHeI \nel) \nHeI \nonumber + \aHeIII \nel \nHeIII \\  & - (\bHeII + \aHeII + \xiHeII) \nel \nHeII \label{eq:HeIIRateEquation} \\ 
    \frac{d \nHeIII}{dt} & = (\gHeII + \gamma_{\HeII} + \bHeII \nel) \nHeII  - \aHeIII \nel \nHeIII . \label{eq:HeIIIRateEquation}
\end{align}
Each of these equations represents the balance between ionizations of species
$\HI$, $\HeI$, and $\HeII$, and recombinations of $\HII$, $\HeII$, and
$\HeIII$. Associating the index $i$ with absorbing species, $i = \HI, \HeI,
\HeII$, and the index $i^{\prime}$ with ions, $i^{\prime} = \HII, \HeII,
\HeIII$, we define $\Gamma_i$ as the photo-ionization rate coefficient,
$\gamma_i$ as the secondary ionization rate coefficient, $\alpha_{i^{\prime}}$
($\xi_{i^{\prime}}$) as the case-B (dielectric) recombination rate
coefficients, $\beta_i$ as the collisional ionization rate coefficients, and
$\nel = \nHII + \nHeII + 2\nHeIII$ as the number density of electrons.

At each time-step, we also solve for the temperature evolution, $dT_k/dt$,
which
is given by
\begin{align}
    \frac{3}{2}\frac{d}{dt}\left(\frac{\kB T_k \ntot}{\mu}\right) & = \fheat  \sum_i n_i \Heat_i - \sum_i \zeta_i n_e n_i - \sum_{i^{\prime}} \eta_{i^{\prime}} n_e n_{i^{\prime}} \nonumber \\ & - \sum_i \psi_i n_e n_i - \omega_{\HeII} n_e \nHeII \label{eq:TemperatureEvolution} 
\end{align}
where $\Heat_i$ is the photo-electric heating rate coefficient (due to
electrons previously bound to species $i$), $\omega_{\HeII}$ is the dielectric
recombination cooling coefficient, and $\zeta_i$, $\eta_{i^{\prime}}$, and
$\psi_i$ are the collisional ionization, recombination, and collisional
excitation cooling coefficients, respectively. The constants in Equation
\ref{eq:TemperatureEvolution} are the total number density of baryons, $\ntot
= n_\mathrm{H} + n_{\mathrm{He}} + n_e$, the mean molecular weight, $\mu$,
Boltzmann's constant, $\kB$, and the fraction of secondary electron energy
deposited as heat, $\fheat$. We use the formulae in Appendix B of
\citet{Fukugita1994} to compute the values of $\alpha_i$, $\beta_i$, $\xi_i$,
$\zeta_i$, $\eta_{i^{\prime}}$, $\psi_i$, and $\omega_{\HeII}$.

% Ionization and Heating Rates
\subsection{Computing the Ionization and Heating Rates}
The most critical aspect of propagating the radiation field in our 1D
simulations is computing the ionization ($\Gamma_i$, $\gamma_{i}$) and heating
($\Heat_i$) rate coefficients accurately. In order to directly relate our
results to fully 3-dimensional radiative transfer calculations we have chosen
to adopt a photon-conserving (PC) algorithm nearly identical to those employed
by several widely used codes, like \textit{C$^2$Ray} \citep{Mellema2006}, and
\textit{Enzo} \citep{Wise2011}. Our code is able to compute $\Gamma_i$,
$\gamma_{i}$, and $\Heat_i$ in a non-photon-conserving (NPC) fashion as well,
to enable comparison with previous 1D work such as \citet{Thomas2008}. The two
formalisms are equivalent in the limit of very optically thin cells, a
condition that can be met easily in 1D calculations but is rarely
computationally feasible in 3D. For NPC methods, if the optical depth of an
individual cell is substantial, the number of ionizations in that cell will
\textit{not} equal the number of photons absorbed for that cell, i.e. photon
number will not be conserved. This problem was remedied by \citet{Abel1999},
who inferred the number of photo-ionizations of species $i$ in a cell from the
radiation incident upon it and its optical depth,
\begin{equation}
    \Delta \tau_{i,\nu} = n_i \sigma_{i,\nu} \Delta r .
\end{equation}    
It is most straightforward to imagine our 1D grid as a collection of
concentric spherical shells, each having thickness $\Delta r$ and volume
$V_{\sh}(r) = 4 \pi [(r + \Delta r)^3 - r^3] / 3$, where $r$ is the distance
between the origin and the inner interface if each shell. The ionization and
heating rates can then be related to the number of absorptions in any given
shell (thus preserving photon number), as
\begin{align}
    \Gamma_i & = A_i \int_{\nu_i}^{\infty} I_{\nu} e^{-\tau_{\nu}} \left(1 - e^{-\Delta \tau_{i,\nu}}\right) \frac{d\nu}{h\nu} \label{eq:PhotoIonizationRate} \\
    \gamma_{ij} & = A_j \int_{\nu_j}^{\infty} \left(\frac{\nu - \nu_j}{\nu_i}\right) I_{\nu} e^{-\tau_{\nu}} \left(1 - e^{-\Delta \tau_{j,\nu}}\right) \frac{d\nu}{h\nu} \label{eq:SecondaryIonizationRate} \\
    \Heat_i & = A_i \int_{\nu_i}^{\infty} (\nu - \nu_i) I_{\nu} e^{-\tau_{\nu}} \left(1 - e^{-\Delta \tau_{i,\nu}}\right) \frac{d\nu}{\nu} , \label{eq:HeatingRate}
\end{align}    
where we have defined the normalization constant $A_i \equiv
L_{\mathrm{bol}}/n_i V_{\sh}(r)$, and denote the ionization threshold energy
for species $i$ as $h\nu_i$.  $I_{\nu}$ represents the SED of radiation sources, and satisfies 
\begin{equation}
\int_{\nu} I_{\nu} d\nu = 1,
\end{equation}
such that $A_i I_{\nu}$ is the number of photons of frequency $\nu$ per atom (of species $i$) in a shell at radius $r$.

@@

% Time-dependent source SEDs
\subsection{Additional Complexity: Time-Dependent Source SEDs}
Our lookup tables will inherit an additional dimension.  Goodie.

% Spectrum Normalization
\subsection{Spectrum Normalization}
For convenience, we generally work with the normalized SED $I_{\nu}$, such that 
\begin{equation}
    L_{\nu} = L_{\mathrm{bol}} I_{\nu} ,
\end{equation}    
and 
\begin{equation}
    \dot{Q}_{\nu} = L_{\nu} / h\nu .
\end{equation}    
However, in \S\ref{sec:Discretization}, we will discuss how \texttt{rt1d} can be used to verify choices of discrete SEDs -- when \texttt{DiscreteSpectrum = 1}, we must be careful about how we normalize a spectrum.

There are two primary ways we can go about normalizing a discrete spectrum consisting of $n_{\nu}$ frequency bins.

The first way is to manually supply the location of each bin (in eV) and the fraction of the bolometric luminosity allocated to that bin via the parameters \texttt{DiscreteSpectrumSED} and \texttt{DiscreteSpectrumRelLum}.  In \citet{Mirocha2012}, we denote these quantities as $h\nu_n$ and $I_n$, respectively, where $n = 1, 2,...n_{\nu}$, and do \textit{not} require the bolometric luminosity to remain conserved.  That is, we allow
\begin{equation}
    \sum_n I_n \neq 1.
\end{equation}    
This is because very high energy photons (depending on the spectrum) may not be absorbed at all over the column density regime or volume of interest.  It is important to note that even if $\sum_n I_n = 1$ exactly, the photon emissivity of a source constructed in this way will \textit{not} be conserved.

For example, when we simulate a $10^5$ K blackbody source emitting $5\times 10^{48}$ ionizing photons per second (as in RT06 problem \#2), the \texttt{RadiationSource} class figures out the bolometric luminosity (show equations here for reference), and assigns a fraction $I_n$ of that to bin $n$.  In doing this, 
\begin{equation}
    L_{\mathrm{bol}} \sum_n I_n / h\nu_n \neq 5 \times 10^{48} \ \mathrm{s}^{-1} .
\end{equation}

The second approach we can use is that used in previous work such as \citet{Aubert2008}, which requires us to define bandpasses, each defined by an average photon energy and effective bound-free absorption cross-section.  The details will be discussed in \S\ref{sec:Discretization}, but for now it is enough to know that the ionizing photon luminosity \textit{will} be preserved with this method, assuming that the bandpasses span the full emission interval of the radiation source of interest, and the bolometric luminosity will \textit{not} be preserved.  That is,
\begin{equation}
    \sum_n \dot{Q}_n h \nu_n \neq L_{\mathrm{bol}} .
\end{equation}

@@

% Lookup Tables
\subsection{Lookup Tables}
How does the dimensionality of our lookup tables depend on input parameters?

@@

%%%
%% SED studies
%%%
\section{Verifying Discrete SEDs with \texttt{rt1d}} \label{sec:Discretization}
In \citet{Mirocha2012}, we used \texttt{rt1d} to verify new, optimally
selected discrete spectral energy distributions for sources of ultraviolet and
X-ray photons. As we optimize more and more SEDs for fully 3D cosmological
radiation-hydro simulations, \texttt{rt1d} remains a quick and easy tool for
verifying that the SEDs we're using are acceptable.

In the literature, discrete SEDs have been constructed using a `multi-group'
approach, which defines new absorption cross sections which represent
spectrum-weighted averages over broad bandpasses,
\begin{equation}
    \tilde{\sigma}_{i,n} \equiv \int_{\nu_n}^{\nu_{n+1}} \sigma_{i,\nu} \frac{I_{\nu}}{h\nu}  d\nu \left(\int_{\nu_n}^{\nu_{n+1}}\frac{I_{\nu}}{h\nu}d\nu \right)^{-1} ,
\end{equation}    
where the interval $(\nu_n, \nu_{n+1})$ represents the bandpass bounds for
discrete photon group $n$.

The emission energies are chosen in a similar way, 
\begin{equation}
    h\nu_n = \int_{h\nu_n}^{h\nu_{n+1}} I_{\nu} d\nu \left(\int_{h\nu_n}^{h\nu_{n+1}} \frac{I_{\nu}}{h\nu} d\nu \right)^{-1} . \label{eq:MeanPhotonEnergy}
\end{equation}    
It's worth keeping in mind that there are different ways to compute the mean
photon energy in each interval. Here, we've calculated by dividing a quantity
proportional to the bolometric energy output by the total number of ionizing
photons. We could also compute the expectation value, median value, or even
the peak of emission (e.g. for a blackbody). Equation
\ref{eq:MeanPhotonEnergy} is most often used in the literature, so we'll stick
to that.
